Early Observations of the Type Ia Supernova \abc:\\ Evidence for Strong Ejecta Mixing and/or Interaction with Diffuse Material

### 2017-08-16 doc/ms.tex
Update the manuscript to reflect the comments from Suhail

\author[0000-0002-0832-2974]{G.~Hosseinzadeh}
\affil{Las Cumbres Observatory, Goleta, CA 93117, USA}
\affil{Physics Department, University of California, Santa Barbara, CA 93106, USA}
\affil{LSSTC Data Science Fellow}

@@

Early observations of Type Ia supernovae (SNe) provide a unique probe of their progenitor systems and explosion physics. Here, we report the intermediate Palomar Transient Factory (iPTF) discovery of an extraordinarily young SN Ia, \abc. By fitting a power law to our early light curve, we infer that first light for the SN, that is when appreciable heating from $^{56}$Ni decay has first diffused to the photosphere, only occurred $0.15\pm_{0.07}^{0.15}$~d before our first detection. In the $\sim$24 hr after discovery, \abc\ rose by $\sim$2 mag, following a near-linear rise in flux for $\sim$4 d. 
% Assuming the photospheric velocity follows a $t^{-0.22}$ power law, the time of explosion for \abc\ is approximately equal to the time of first light.
Strong \ion{C}{2} absorption is detected in the early spectra of \abc, before disappearing after $\sim$7~d. Unlike SN\,2011fe, the $(B-V)_0$ colors of \abc\ are blue and nearly constant in the days after explosion. We show that our early observations of \abc\ cannot be explained by either SN shock breakout and the associated, subsequent cooling, or the SN ejecta colliding with a stellar companion. Instead, we argue that the early characteristics of \abc, including: (i) the rapid, near-linear rise, (ii) the non-evolving blue colors, and (iii) the strong absorption from ionized carbon, are the result of either vigorous mixing of radioactive $^{56}$Ni in the SN ejecta, or ejecta interaction with diffuse material, or a combination of the two. In the next few years, dozens of very young \textit{normal} SNe Ia will be discovered, and observations similar to those presented here will constrain the white dwarf explosion mechanism.

@@

After establishing \abc\ as a normal SN Ia, we use the latest
calibration \citep{2014A&A...568A..22B} of the Phillips relation
\citep{1993ApJ...413L.105P} using $m^*_{B}$, $x_1$ and $c$ to derive a distance modulus $\mu = 34.89 \pm 0.10 \,\textrm{mag}$ to the SN, provided that the host galaxy of \abc\ has a stellar mass $< 10^{10}\sm$. A more massive host galaxy would result in a larger inferred distance modulus that is nevertheless consistent within the uncertainties. For the following analysis we adopt a distance modulus $\mu = 34.89 \pm 0.10 \,\textrm{mag}$ for \abc.

@@

The potential dark phase following shock breakout, where SNe Ia are not luminous because the energy deposition from radioactive decay has not had sufficient time to diffuse to the photosphere, means that the explosion time $t_\mathrm{exp} \ne t_0$ in some cases. \citet{2014ApJ...784...85P} suggest that measurements of the photospheric velocity be used to determine $t_\mathrm{exp}$ given that the ejecta begin expanding from the moment of explosion.  Assuming a constant opacity in the ejecta, \citeauthor{2014ApJ...784...85P} 
find that the photospheric velocity evolves as
$v_{ph}\propto(t-t_\mathrm{exp})^{-0.22}$. Numerical experiments by \citet{2016ApJ...826...96P} find that the consatnt-opacity assumption strongly depends on the amount of $^{56}$Ni mixing in the SN ejecta (for a more detailed discussion of $^{56}$Ni mixing see \S\ref{sec:Ni_mixing}). As a result, the adoption of a $t^{-0.22}$ power-law model may not be valid for all SNe Ia. Nevertheless, we proceed on the assumption that \abc\ experienced strong $^{56}$Ni mixing, corresponding to the models that are best approximated as a $t^{-0.22}$ power law. We do this in part to compare with previous studies, though we caution that the inferred value of $t_{exp}$ is subject to uncertainties related to ejecta mixing.

@@

To model SN\,2011fe, \citet{2014MNRAS.441..532D} recently examined pulsational delayed-detonation (PDD) models as an explanation for some SNe Ia. Briefly, PDD models differ from ``standard'' delayed detonation (DD) models in that the initial deflagration causes the WD to expand resulting in the release of some unbound material. As the bound material contracts, eventually a subsequent detonation occurs.\footnote{\citet{2014MNRAS.441..532D} note that the deflagration and detonation in their PDD models are artificially triggered.} An important consequence of this progression for PDD models is that they naturally produce material that avoids burning, unlike DD models that typically leave no unburnt material. This results in significantly more carbon in the outer layers of the SN ejecta \citep{2014MNRAS.441..532D}.

>>>>

\author[0000-0002-0832-2974]{G.~Hosseinzadeh}
\affil{Las Cumbres Observatory, Goleta, CA 93117, USA}
\affil{Physics Department, University of California, Santa Barbara, CA 93106, USA}

@@

Early observations of Type Ia supernovae (SNe) provide a unique probe of their progenitor systems and explosion physics. Here, we report the intermediate Palomar Transient Factory (iPTF) discovery of an extraordinarily young SN Ia, \abc. By fitting a power law to our early light curve, we infer that first light for the SN, that is when appreciable heating from $^{56}$Ni decay has first diffused to the photosphere, only occurred $0.15\pm_{0.07}^{0.15}$~d before our first detection. In the $\sim$24 hr after discovery, \abc\ rose by $\sim$2 mag, following a near-linear rise in flux for $\gtrsim$3 d. 
Strong \ion{C}{2} absorption is detected in the early spectra of \abc, before disappearing after $\sim$7~d. Unlike the extensively-observed Type Ia SN\,2011fe, the $(B-V)_0$ colors of \abc\ are blue and nearly constant in the days after explosion. We show that our early observations of \abc\ cannot be explained by either SN shock breakout and the associated, subsequent cooling, or the SN ejecta colliding with a stellar companion. Instead, we argue that the early characteristics of \abc, including: (i) the rapid, near-linear rise, (ii) the non-evolving blue colors, and (iii) the strong absorption from ionized carbon, are the result of either vigorous mixing of radioactive $^{56}$Ni in the SN ejecta, or ejecta interaction with diffuse material, or a combination of the two. In the next few years, dozens of very young \textit{normal} SNe Ia will be discovered, and observations similar to those presented here will constrain the white dwarf explosion mechanism.

@@

After establishing \abc\ as a normal SN Ia, we use the latest
calibration \citep{2014A&A...568A..22B} of the Phillips relation
\citep{1993ApJ...413L.105P} using $m^*_{B}$, $x_1$ and $c$ to derive a distance modulus $\mu = 34.89 \pm 0.10 \,\textrm{mag}$ to the SN, provided that the host galaxy of \abc\ has a stellar mass $< 10^{10}\sm$. A more massive host galaxy would result in a larger inferred distance modulus that is nevertheless consistent within the uncertainties. For the following analysis we adopt a distance modulus $\mu = 34.89 \pm 0.10 \,\textrm{mag}$ for \abc.\footnote{This $\mu$ is consistent with the $z_\mathrm{SN}$, $H_0 = 73 \, \mathrm{km \, s^{-1} \, Mpc}^{-1}$, and Virgo-infall-corrected distance \citep{2000ApJ...529..786M}.}

@@

The potential dark phase following shock breakout, where SNe Ia are not luminous because the energy deposition from radioactive decay has not had sufficient time to diffuse to the photosphere, means that the explosion time $t_\mathrm{exp} \ne t_0$ in some cases. \citet{2014ApJ...784...85P} suggest that measurements of the photospheric velocity be used to determine $t_\mathrm{exp}$ given that the ejecta begin expanding from the moment of explosion.  Assuming a constant opacity in the ejecta, \citeauthor{2014ApJ...784...85P} 
find that the photospheric velocity evolves as
$v_\mathrm{ph}\propto(t-t_\mathrm{exp})^{-0.22}$. Numerical experiments by \citet{2016ApJ...826...96P} find that the consatnt-opacity assumption strongly depends on the amount of $^{56}$Ni mixing in the SN ejecta (for a more detailed discussion of $^{56}$Ni mixing see \S\ref{sec:Ni_mixing}). As a result, the adoption of a $t^{-0.22}$ power-law model may not be valid for all SNe Ia. Nevertheless, we proceed on the assumption that \abc\ experienced strong $^{56}$Ni mixing, corresponding to the models that are best approximated as a $t^{-0.22}$ power law. We do this in part to compare with previous studies, though we caution that the inferred value of $t_{exp}$ is subject to uncertainties related to ejecta mixing.

@@

To model SN\,2011fe, \citet{2014MNRAS.441..532D} recently examined pulsational delayed-detonation (PDD) models as an explanation for some SNe Ia. Briefly, PDD models differ from ``standard'' delayed detonation (DD) models in that the expansion of the WD during the initial deflagration phase leads to the release of unbound material. Following this pulsation, the bound material contracts eventually triggering a subsequent detonation.\footnote{\citet{2014MNRAS.441..532D} note that the deflagration and detonation in their PDD models are artificially triggered.} An important consequence of this progression for PDD models is that the unbound material expands and avoids burning, unlike DD models that typically leave no unburnt material. This results in significantly more carbon in the outer layers of the SN ejecta \citep{2014MNRAS.441..532D}.
